### 2016-07-04 store-consistency.tex

In practice, relational databases resolve such conflicts by
prioritizing weak isolation (thus, concurrency and performance) over
strong consistency, so the execution traces do not necessarily satisfy
{\sc sc}. In particular, visibility constraints imposed by {\sc sc}
are violated iff they are found to be in conflict with the constraints
imposed by a transaction's isolation level. In the context of the
aforementioned example, $\eta_1$ is not made visible to $\eta_2$
because doing so would violate $\I_c$. However, if $\I_c$ is redefined
to \emph{true}, then the store makes $\eta_1$ visible to $\eta_2$ to
honor its consistency commitment\footnote{The term \emph{recency
commitment}~\cite{bailishat} is often used in practice to capture the
best-effort nature of {\sc sc}.}. We generalize this approach to any
$\I_s$ and $\I_c$ by defining a \emph{maximum visibility principle} to
determine an acceptable weakening of $\I_s$ in case of a conflict with
$\I_c$.  The principle requires the weakened consistency guarantee
($\I_s'$) of the store to enforce all visibility relationships imposed
by the actual consistency guarantee ($\I_s$), unless enforcing such a
relationship violates $\I_c$.
% The formal definition of the principle is relegated
% to the supplementary in the interest of space, but it can be
% understood in the context of an {\sc sc} store, where it weakens the
% store invariant to the following:
Formally:
\begin{definition}
$\I_s' : \E \rightarrow \Prop$ is said to be a maximum visibility
weakening of $\I_s : \E \rightarrow \Prop$ if and only if:
\begin{itemize}
  \item $\I_s'$ is weaker than $\I_s$: 
      $\forall\E.~ \I_s(\E) \Rightarrow \I_s'(\E)$, and
  \item In every trace $\E$ that satisfies $\I_c$, and for every pair
  of effects $\eta_1$ and $\eta_2$ in $\E$, if $\I_s(\E)$ requires
  $\eta_1$ to be visible to $\eta_2$, then so does $\I_s'(\E)$ unless
  extending $\E$ with $\visZ(\eta_1,\eta_2)$ violates
  $\I_c$:
  \begin{smathpar}
  \begin{array}{l}
  \forall\E,\eta_1,\eta_2.~ \I_c(\E) \Rightarrow (\I_s(\E)
    \Rightarrow \underE{\eta_1 \visar \eta_2}) \Rightarrow \\
    \hspace*{0.5in}(\I_s'(\E) \Rightarrow \underE{\eta_1 \visar
    \eta_2} \disj \neg\I_c(\E\,\cup\,(\emptyset,\{(\eta_1,\eta_2)\})))
  \end{array}
  \end{smathpar}
\end{itemize}
\end{definition}
Applying this principle, we can weaken {\sc sc} to obtain the
following store trace invariant ($\I_s$) for an {\sc sc} store whose
isolation constraints are captured by $\I_c$:
\begin{smathpar}
\begin{array}{lcl}
\I_s(\E) & = & \forall \eta_1,\eta_2.\, \{\eta_1,\eta_2\},
    \subseteq \E.\A \conj \id(\eta_1) <
    \id(\eta_2) \\
    & & \hspace*{0.5in} \Rightarrow 
      \underE{\eta_1 \visar \eta_2} \disj \neg\I_c(\E
    \cup (\emptyset,\{(\eta_1,\eta_2)\}))\\
\end{array}
\end{smathpar}
% GK : Please talk to me before changing the substance of the following paragraph.
$\I_s$ requires a trace $\E$ to satisfy the visibility constraints of
{\sc sc} except in cases where they are in conflict with $\I_c$.
Instantiating the parameter $\I$ with $\I_s \wedge \I_c$ in
Fig.~\ref{fig:txnimp} results in an operational semantics that
describes {\sc sc} stores with on-demand weak isolation, such as
off-the-shelf relational databases. Similar approaches can be employed
to capture the semantics of other kinds of stores, such as a causally
consistent ({\sc cc}) data store, as specific instantiations of our
operational semantics\footnote{Details of the {\sc cc} instantiation
are relegated to the supplementary in the interest of space.}.

@@

Decorated parts of \C{Wd1} relevant to the discussion are shown in
Fig.~\ref{fig:wd1-decorated}.  \C{Wd2} is similar and not
shown\footnote{Supplementary contains fully decorated program.}.
Temporary name \C{v3} has been introduced to capture the result of the
RHS expression \C{C-a1}.  Assertions implicitly refer to the current
execution ($\E$), just as Hoare triples implicitly refer to the
current state.  The context for propositions is also the implicit,
i.e., we write $\psi$ instead of $\underE{\psi}$. The proposition \C{k
$\ge$ a1+a2} remains an invariant, hence elided. The pre-condition
$P_1$ of \C{Wd1} accounts for the possibility of \C{Wd2} committing
before \C{Wd1}, writing \C{k-a2} to \C{C}. Since neither \C{Wd1} nor
\C{Wd2} are committed at the beginning, the pre-condition $P$ of the
program is extended with $\neg\committed(\C{Wd1}) \wedge
\neg\committed(\C{Wd2})$, from which $P_1$ follows. The post-condition
$Q_1$ of \C{Wd1} asserts different values for \C{C} depending on
whether or not \C{Wd2} has already committed.  It also asserts that
\C{Wd1} has committed. The post-condition $Q_2$ of \C{Wd2} (not shown)
is similar, but with the roles of \C{Wd1} and \C{Wd2} exchanged. The
post-condition $Q$ follows from $Q_1 \wedge Q_2$, hence valid. It
remains to show that \C{Wd1} satisfies its specification (\C{Wd2} must
also satisfy its spec, but the proof is similar).

>>>>

In practice, relational databases resolve such conflicts by
prioritizing weak isolation (thus, concurrency and performance) over
strong consistency, so the execution traces do not necessarily satisfy
{\sc sc}. In particular, visibility constraints imposed by {\sc sc}
are violated iff they are found to be in conflict with the constraints
imposed by a transaction's isolation level. In the context of the
aforementioned example, $\eta_1$ is not made visible to $\eta_2$
because doing so would violate $\I_c$. However, if $\I_c$ is redefined
to \emph{true}, then the store makes $\eta_1$ visible to $\eta_2$ to
honor its consistency commitment\footnote{The term \emph{recency
commitment}~\cite{bailishat} is often used in practice to capture the
best-effort nature of {\sc sc}.}. We generalize this approach to any
$\I_s$ and $\I_c$ by defining a \emph{maximum visibility principle} to
determine an acceptable weakening of $\I_s$ in case of a conflict with
$\I_c$.  The principle requires the weakened consistency guarantee
($\I_s'$) of the store to enforce all visibility relationships imposed
by the actual consistency guarantee ($\I_s$), unless enforcing such a
relationship violates $\I_c$.
% The formal definition of the principle is relegated
% to the supplementary in the interest of space, but it can be
% understood in the context of an {\sc sc} store, where it weakens the
% store invariant to the following:
Formally:
\begin{definition}
$\I_s' : \E \rightarrow \Prop$ is said to be a maximum visibility
weakening of $\I_s : \E \rightarrow \Prop$ if and only if:
\begin{itemize}
  \item $\I_s'$ is weaker than $\I_s$: 
      $\forall\E.~ \I_s(\E) \Rightarrow \I_s'(\E)$, and
  \item In every trace $\E$ that satisfies $\I_c$, and for every pair
  of effects $\eta_1$ and $\eta_2$ in $\E$, if $\I_s(\E)$ requires
  $\eta_1$ to be visible to $\eta_2$, then so does $\I_s'(\E)$ unless
  extending $\E$ with $\visZ(\eta_1,\eta_2)$ violates
  $\I_c$:
  \begin{smathpar}
  \begin{array}{l}
  \forall\E,\eta_1,\eta_2.~ \I_c(\E) \Rightarrow (\I_s(\E)
    \Rightarrow \underE{\eta_1 \visar \eta_2}) \Rightarrow \\
    \hspace*{0.5in}(\I_s'(\E) \Rightarrow \underE{\eta_1 \visar
    \eta_2} \disj \neg\I_c(\E\,\cup\,(\emptyset,\{(\eta_1,\eta_2)\})))
  \end{array}
  \end{smathpar}
\end{itemize}
\end{definition}
Applying this principle, we can weaken {\sc sc} to obtain the
following store trace invariant ($\I_s$) for an {\sc sc} store whose
isolation constraints are captured by $\I_c$:
\begin{smathpar}
\begin{array}{lcl}
\I_s(\E) & = & \forall \eta_1,\eta_2.\, \{\eta_1,\eta_2\},
    \subseteq \E.\A \conj \id(\eta_1) <
    \id(\eta_2) \\
    & & \hspace*{0.5in} \Rightarrow 
      \underE{\eta_1 \visar \eta_2} \disj \neg\I_c(\E
    \cup (\emptyset,\{(\eta_1,\eta_2)\}))\\
\end{array}
\end{smathpar}
% GK : Please talk to me before changing the substance of the following paragraph.
$\I_s$ requires a trace $\E$ to satisfy the visibility constraints of
{\sc sc} except in cases where they are in conflict with $\I_c$.
Instantiating the parameter $\I$ with $\I_s \wedge \I_c$ in
Fig.~\ref{fig:txnimp} results in an operational semantics that
describes {\sc sc} stores with on-demand weak isolation, such as
provided by off-the-shelf relational databases. Similar approaches can be employed
to capture the semantics of other kinds of stores, such as a causally
consistent ({\sc cc}) data store, as specific instantiations of our
operational semantics\footnote{Details of the {\sc cc} instantiation
are given in the supplementary material.}.

@@

Decorated parts of \C{Wd1} relevant to the discussion are shown in
Fig.~\ref{fig:wd1-decorated}.  \C{Wd2} is similar and not
shown\footnote{The supplementary material shows the fully decorated program.}.
Temporary name \C{v3} has been introduced to capture the result of the
RHS expression \C{C-a1}.  Assertions implicitly refer to the current
execution ($\E$), just as Hoare triples implicitly refer to the
current state.  The context for propositions is also the implicit,
i.e., we write $\psi$ instead of $\underE{\psi}$. The proposition \C{k
$\ge$ a1+a2} remains an invariant, hence elided. The pre-condition
$P_1$ of \C{Wd1} accounts for the possibility of \C{Wd2} committing
before \C{Wd1}, writing \C{k-a2} to \C{C}. Since neither \C{Wd1} nor
\C{Wd2} are committed at the beginning, the pre-condition $P$ of the
program is extended with $\neg\committed(\C{Wd1}) \wedge
\neg\committed(\C{Wd2})$, from which $P_1$ follows. The post-condition
$Q_1$ of \C{Wd1} asserts different values for \C{C} depending on
whether or not \C{Wd2} has already committed.  It also asserts that
\C{Wd1} has committed. The post-condition $Q_2$ of \C{Wd2} (not shown)
is similar, but with the roles of \C{Wd1} and \C{Wd2} exchanged. The
post-condition $Q$ follows from $Q_1 \wedge Q_2$, hence valid. It
remains to show that \C{Wd1} satisfies its specification (\C{Wd2} must
also satisfy its spec, but the proof is similar).
